### 2019-12-12 ch10_discussion/content.tex
update

\section{Outlook}
The types of computational analyses presented in Section~\ref{sec:comptools} (Clustering, DE, DR, TI, NI) have all become routine methodologies for analysing and interpreting single-cell omics data. With projects like the Human Cell Atlas\cite{humancellatlasconsortium_humancellatlas_2018} soon to be churning out millions of single-cell profiles on a regular basis, research in high-throughput computational tools to analyse these data in an unsupervised approach becomes increasingly relevant.

@@

Trajectory alignment has been used to compare gene expression kinetics resulting from different biological processes\cite{cacchiarelli_aligningsinglecelldevelopmental_2018}, to compare human, chimpanzee and macaque neuronal development\cite{kanton_organoidsinglecellgenomic_2019}, to find differences in gene regulation in the presence of certain growth factors\cite{mcfaline-figueroa_pooledsinglecellgenetic_2019}, and to compare human and mouse embryogenesis\cite{alpert_alignmentsinglecelltrajectories_2018}.

>>>>

\section{Outlook}
The types of computational analyses presented in Section~\ref{sec:comptools} (Clustering, DE, DR, TI, NI) have all become routine methodologies for analysing and interpreting single-cell omics data. With projects like the Human Cell Atlas \cite{humancellatlasconsortium_humancellatlas_2018} soon to be churning out millions of single-cell profiles on a regular basis, research in high-throughput computational tools to analyse these data in an unsupervised approach becomes increasingly relevant.

@@

Trajectory alignment has been used to compare gene expression kinetics resulting from different biological processes \cite{cacchiarelli_aligningsinglecelldevelopmental_2018}, to compare human, chimpanzee and macaque neuronal development \cite{kanton_organoidsinglecellgenomic_2019}, to find differences in gene regulation in the presence of certain growth factors \cite{mcfaline-figueroa_pooledsinglecellgenetic_2019}, and to compare human and mouse embryogenesis \cite{alpert_alignmentsinglecelltrajectories_2018}.

### 2019-12-12 ch1_introduction/content.tex
update

\newpage 
\section{Hippocratic oath for method developers}
Before setting out on this perilous journey, I write a Hippocratic oath 
which I believe all developers of computational tools should swear by 
before haphazardly creating new tools.

\begin{figure}[htb!]
	\centering
	\begin{tcolorbox}
		\begin{itemize}
			\item In developing a new computational tool, I swear to fulfil, to the best of my ability, the following items:
			\item I will write software that works reliably but fails gracefully when it does not.
			\item I will assert that my method produces accurate and reproducible results.
			\item I will validate the performance of my tool with previously unseen data from third-party sources.
			\item I will use suitable quantitative performance metrics. If possible, I use multiple metrics to avoid overfitting.
			\item I will compare the performance to relevant methods.
			\item I will report positive achievements of my method, but also be critical of it, and discuss areas in which it performs suboptimally.
			\item I will write unit tests to ensure that each component in my tool works as intended. If the tool produces unexpected errors when tested on various edge cases, my tool will be perceived as dysfunctional by others.
			\item I will follow reproducible research best practices, by making code and data publicly available.
		\end{itemize}
	\end{tcolorbox}
	\caption{
		\textbf{A hippocratic oath for computational method developers.} Inspired by Laplante\cite{laplante_firstnoharm_2004} and lists of guidelines for benchmarking\cite{norel_selfassessmenttrapcan_2011,weber_essentialguidelinescomputational_2019}
	}
	\label{fig:hippocratic_oath}
\end{figure}

>>>>

\newpage 
\section{Hippocratic oath for method developers}
Before setting out on this perilous journey, I write a Hippocratic oath which I believe all developers of computational tools should swear by before haphazardly developing and publishing new software tools.

\begin{figure}[htb!]
	\centering
	\begin{tcolorbox}
		\begin{itemize}
			\item In developing a new computational tool, I swear to fulfil, to the best of my ability, the following items:
			\item I will write software that works reliably but fails gracefully when it does not.
			\item I will assert that my method produces accurate and reproducible results.
			\item I will validate the performance of my tool with previously unseen data from third-party sources.
			\item I will use suitable quantitative performance metrics. If possible, I use multiple metrics to avoid overfitting.
			\item I will compare the performance to relevant methods.
			\item I will report positive achievements of my method, but also be critical of it, and discuss areas in which it performs sub-optimally
			\item I will write unit tests to ensure that each component in my tool works as intended. If the tool produces unexpected errors when tested on various edge cases, my tool will be perceived as dysfunctional by others.
			\item I will follow reproducible research best practices, by making code and data publicly available.
		\end{itemize}
	\end{tcolorbox}
	\caption{
		\textbf{A Hippocratic oath for computational method developers.} Inspired by Laplante\cite{laplante_firstnoharm_2004} and lists of guidelines for benchmarking\cite{norel_selfassessmenttrapcan_2011,weber_essentialguidelinescomputational_2019}
	}
	\label{fig:hippocratic_oath}
\end{figure}

### 2019-12-12 ch3_dynbenchmark/content.tex
update

\section{Introduction}
Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}. Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

@@

We used the Github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores. About one-third of the authors responded and we improved the wrappers based on their feedback. These discussions can be viewed on Github: \url{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}

@@

\begin{itemize}
	\item \textbf{Start cells}: the identity of one or more start cells. For both real and synthetic data, a cell was chosen that was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
	\item \textbf{End cells}: the identity of one or more end cells. This is similar to the start cells, but now for every state with only incoming edges.
	\item \textbf{No. of end states}: number of terminal states, i.e., the number of milestones with only incoming edges.
	\item \textbf{Grouping}: for each cell a label showing which state/cluster/branch it belongs to. For real data, the states were from the gold/silver standard. For synthetic data, each milestone was seen as one group and cells were assigned to their closest milestone.
	\item \textbf{No. of branches}: number of branches/intermediate states. For real data, this was the number of states in the gold/silver standard. For synthetic data, this was the number of milestones.
	\item \textbf{Discrete time course}: for each cell a time point from which it was sampled. If available, this was directly extracted from the reference trajectory; otherwise the geodesic distance from the root milestone was used. For synthetic data, the simulation time was uniformily discretized into four timepoints.
	\item \textbf{Continuous time course}: for each cell a time point from which it was sampled. For real data, this was equal to the discrete time course. For synthetic data, we used the internal simulation time of each simulator.
\end{itemize}

@@

The dyngen (Chapter~\ref{chap:dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. We simulated a total of 110 datasets, with 11 different topologies.

@@

Using the simulate\_lineage function, we simulated the lineage expression, with parameters $a \sim U(0.01, 0.1)$, $\textit{branch-tol}_{\textit{intra}} \sim U(0, 0.9)$ and $\textit{branch-tol}_{\textit{inter}} \sim U(0, 0.9)$. These parameter distributions were chosen very broad so as to make sure both easy and difficult datasets are simulated. After simulating base gene expression with simulate\_base\_gene\_exp, we used the sample\_density function to finally simulate expression values of a number of cells (the same as from the reference real dataset), with $\alpha \sim \textit{Lognormal}$ ($\mu = 0.3$ and $\sigma = 1.5$) and $\beta \sim \textit{Lognormal}$ ($\mu = 2$ and $\sigma = 1.5$). Each of these parameters were centered around the default values of PROSSTT, but with enough variability to ensure a varied set of datasets.

>>>>

\section{Introduction}
Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. Such dynamic processes can be modelled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}. Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

@@

We used the GitHub issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores. About one-third of the authors responded and we improved the wrappers based on their feedback. These discussions can be viewed on GitHub: \url{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}

@@

\begin{itemize}
	\item \textbf{Start cells}: the identity of one or more start cells. For both real and synthetic data, a cell was chosen that was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
	\item \textbf{End cells}: the identity of one or more end cells. This is similar to the start cells, but now for every state with only incoming edges.
	\item \textbf{No. of end states}: number of terminal states, i.e., the number of milestones with only incoming edges.
	\item \textbf{Grouping}: for each cell a label showing which state/cluster/branch it belongs to. For real data, the states were from the gold/silver standard. For synthetic data, each milestone was seen as one group and cells were assigned to their closest milestone.
	\item \textbf{No. of branches}: number of branches/intermediate states. For real data, this was the number of states in the gold/silver standard. For synthetic data, this was the number of milestones.
	\item \textbf{Discrete time course}: for each cell a time point from which it was sampled. If available, this was directly extracted from the reference trajectory; otherwise the geodesic distance from the root milestone was used. For synthetic data, the simulation time was uniformly discretised into four timepoints.
	\item \textbf{Continuous time course}: for each cell a time point from which it was sampled. For real data, this was equal to the discrete time course. For synthetic data, we used the internal simulation time of each simulator.
\end{itemize}

@@

The dyngen (Chapter~\ref{chap:dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendible We simulated a total of 110 datasets, with 11 different topologies.

@@

Using the simulate\_lineage function, we simulated the lineage expression, with parameters $a \sim U(0.01, 0.1)$, $\textit{branch-tol}_{\textit{intra}} \sim U(0, 0.9)$ and $\textit{branch-tol}_{\textit{inter}} \sim U(0, 0.9)$. These parameter distributions were chosen very broad so as to make sure both easy and difficult datasets are simulated. After simulating base gene expression with simulate\_base\_gene\_exp, we used the sample\_density function to finally simulate expression values of a number of cells (the same as from the reference real dataset), with $\alpha \sim \textit{Lognormal}$ ($\mu = 0.3$ and $\sigma = 1.5$) and $\beta \sim \textit{Lognormal}$ ($\mu = 2$ and $\sigma = 1.5$). Each of these parameters were centred around the default values of PROSSTT, but with enough variability to ensure a varied set of datasets.
